Parallel Tractability of Ontology\\ Materialization: Technique and Practice

### 2018-02-27 parallel-tractability-J-v9/background.tex
finished checking the appendix, merges changes into rel. work and conclusions
fixed the point of \sqsubseteq_*
final polishing

In what follows, for an ontology $\mathcal{O}=\langle\mathcal{T},\mathcal{R},\mathcal{A}\rangle$,
we use $P=\langle R, \textbf{I}\rangle$ to represent the corresponding datalog program
where $R$ is the set of rules obtained by transforming the axioms in
$\mathcal{T}$ and $\mathcal{R}$ and
$\textbf{I}$ is the set of facts that are directly copied from the assertions in $\mathcal{A}$.
Further, we use $R_1\sqsubseteq_{*}R_2$ to denote the smallest transitive reflexive relation
between roles such that $R_1\sqsubseteq R_2\in\mathcal{R}$ implies $R_1\sqsubseteq_{*}R_2$
and $R_1^-\sqsubseteq_{*}R_2^-$. In this paper, we also use the
notion of \emph{simple role}, which was initially proposed to restrict the
usage of highly expressive ontology languages \cite{HorrocksS04}.
Specifically, a role $S\in\textbf{R}$ is \emph{simple} if, (1) it has no subrole (including $S$)
occurring on the right-hand side of axioms of the forms (R3) and (R4); (2) $S^-$ is simple.

Based on the above representations, ontology materialization
corresponds to the evaluation of datalog programs.
Specifically, given a datalog program $P=\langle R, \textbf{I}\rangle$,
let $T_R(\textbf{I})=\{H\theta|\forall B_1,\ldots,B_n\rightarrow H\in R,
B_i\theta\in\textbf{I}, 1\leq i\leq n \}$,
where $\theta$ is some substitution; further let $T_R^{0}(\textbf{I})=\textbf{I}$ and
 $T_R^{i}(\textbf{I})=T_R^{i-1}(\textbf{I})\cup T_R(T_R^{i-1}(\textbf{I}))$ for each $i>0$.
The smallest integer $n$ such that $T_R^{n}(\textbf{I})= T_R^{n+1}(\textbf{I})$ is called \emph{stage},
and \emph{materialization} refers to the computation of $T_R^{n}(\textbf{I})$ with respect to $R$ and \textbf{I}.
$T_R^{n}(\textbf{I})$ is also called the \emph{fixpoint} and denoted by $T_R^{\omega}(\textbf{I})$.
We say that an atom $A$ is \emph{derivable} or can be \emph{derived} with respect
to the datalog program $P=\langle R, \textbf{I}\rangle$ if $A\in T_R^{\omega}(\textbf{I})$.
In this paper, we consider the data complexity of materialization, i.e., we \emph{assume that the rule set $R$ is fixed}
for any class of datalog programs.

>>>>

In what follows, for an ontology $\mathcal{O}=\langle\mathcal{T},\mathcal{R},\mathcal{A}\rangle$,
we use $P=\langle R, \textbf{I}\rangle$ to represent the corresponding datalog program
where $R$ is the set of rules obtained by transforming the axioms in
$\mathcal{T}$ and $\mathcal{R}$ and
$\textbf{I}$ is the set of facts that are directly copied from the assertions in $\mathcal{A}$.
Further, we use $R_1\sqsubseteq_{*}R_2$ to denote the smallest transitive reflexive relation
between roles such that $R_1\sqsubseteq R_2\in\mathcal{R}$ implies $R_1\sqsubseteq_{*}R_2$
and $R_1^-\sqsubseteq_{*}R_2^-$. We omit the reference to
$\mathcal{R}$ if it is clear from the context. In this paper, we also use the
notion of \emph{simple role}, which was initially proposed to restrict the
usage of highly expressive ontology languages \cite{HorrocksS04}.
Specifically, a role $S\in\textbf{R}$ is \emph{simple} if, (1) it has no subrole (including $S$)
occurring on the right-hand side of axioms of the forms (R3) and (R4); (2) $S^-$ is simple.

Based on the above representations, ontology materialization
corresponds to the evaluation of datalog programs.
Specifically, given a datalog program $P=\langle R, \textbf{I}\rangle$,
we set\\[.5\baselineskip]
\mbox{}\hfill $T_R(\textbf{I})=\{H\theta \mid \forall B_1,\ldots,B_n\rightarrow H\in R,
B_i\theta\in\textbf{I}, 1\leq i\leq n \}$,\hfill\mbox{}\\[.5\baselineskip]
where $\theta$ is some substitution; further let $T_R^{0}(\textbf{I})=\textbf{I}$ and
 $T_R^{i}(\textbf{I})=T_R^{i-1}(\textbf{I})\cup T_R(T_R^{i-1}(\textbf{I}))$ for each $i>0$.
The smallest integer $n$ such that $T_R^{n}(\textbf{I})= T_R^{n+1}(\textbf{I})$ is called \emph{stage},
and \emph{materialization} refers to the computation of $T_R^{n}(\textbf{I})$ with respect to $R$ and \textbf{I}.
$T_R^{n}(\textbf{I})$ is also called the \emph{fixpoint} and denoted by $T_R^{\omega}(\textbf{I})$.
We say that an atom $A$ is \emph{derivable} or can be \emph{derived} with respect
to the datalog program $P=\langle R, \textbf{I}\rangle$ if $A\in T_R^{\omega}(\textbf{I})$.
In this paper, we consider the data complexity of materialization, i.e., we \emph{assume that the rule set $R$ is fixed}
for any class of datalog programs.

### 2018-02-27 parallel-tractability-J-v9/evaluation.tex
final polishing

\subsection{Parallel Tractability of the LUBM Datasets and YAGO Ontologies}
\label{sec:lubm-yago}

\textbf{LUBM}. In the Semantic Web community, LUBM
(The Lehigh University Benchmark) is proposed to

@@

Figure~\ref{fig:diffdepths} shows that MG-depths indeed determines
the reasoning times.
For the five datasets ptu-$4m$-$i$, $i\in\{1,2,4,8,16\}$, the experimental
results show an obvious downtrend of the reasoning times with declining MG-depths for
both ParallelDHL and RDFox (see Figure~\ref{fig:diffdepths}(left)).
The downtrend of reasoning times also exists
when handling the datasets nptu-$4m$-$i$, $i\in\{1,2,4,8,16\}$ (see Figure~\ref{fig:diffdepths}(right)).
Although the PTU and NPTU datasets are unrealistic in practice, these experiments indeed
verify that MG-depths determines the reasoning time in parallel
considering that the input ontology sizes are close.

@@

We now discuss the issue of parallel tractability based on the experimental
results of ptu-$4m$-1 and nptu-$4m$-1.
For the dataset ptu-$4m$-1, the trends of reasoning times are different
between ParallelDHL and RDFox.
It can be checked from Table~\ref{tab:speedup} that
RDFox does not benefit from the allocation of several threads for ptu-$4m$-1;
the speedups under different numbers of threads stay around 1.
This means that reasoning times are not reduced with an increasing number of threads.
ParallelDHL shows a better speedup effect. With only one allocated thread, ParallelDHL finishes
the materialization of ptu-$4m$-1 in $135.79$ seconds (Table~\ref{tab:speedup}). When four and more threads
are allocated, the reasoning performance is improved significantly.
The maximal speedup is $5.23$ using $24$ threads.
Specially, with $16$ or more threads, ParallelDHL
requires less time to finish the materialization compared to RDFox.
From Figure~\ref{fig:diffthreads}(left), there is an obvious downtrend of the reasoning times
from one thread to $24$ threads.
The main reason for the difference between ParallelDHL and RDFox
is that the PTU ontology is tractable in parallel and ParallelDHL
is optimized based on SWD paths. The PTU ontology follows the
simple-concept restriction. Thus, there is no twisting path in
the materialization graph of ptu-$4m$-1 although all articles are involved
in one reference chain. This allows for computing the relation $S_{\textit{\!\tiny
    rch}}$ only once in the second iteration of \ref{alg3:updateG} of Algorithm~$\mathsf{A}_{\text{prc}}$
when handling ptu-$4m$-1 (referring to the analysis for Example~\ref{exp:simpleC}).
It can also be checked that all the axioms of the form \texttt{CollegeArticle}$(a_i)$ occur in
an SWD path. Thus, the optimization based on SWD paths works when handling ptu-$4m$-1.
For RDFox, since it is not optimized specially for this case, it performs the task of
materialization over ptu-$4m$-1 similarly to that over nptu-$4m$-1;
specifically, the axioms of the form \texttt{CollegeArticle}$(a_i)$ are derived sequentially.
From Figure~\ref{fig:diffdepths}(left), we can see that the optimization used in ParallelDHL
also leads to a better performance compared to
RDFox when handling ptu-$4m$-1, ptu-$4m$-2 and ptu-$4m$-4 using $24$ threads.

>>>>

\subsection{Parallel Tractability of LUBM Datasets and YAGO}
\label{sec:lubm-yago}

\textbf{LUBM}. In the Semantic Web community, 
the Lehigh University Benchmark (LUBM) is proposed to

@@

Figure~\ref{fig:diffdepths} shows that MG-depths indeed determines
the reasoning times.
For the five datasets ptu-$4m$-$i$, $i\in\{1,2,4,8,16\}$, the experimental
results show an obvious downtrend of the reasoning times with declining MG-depths for
both ParallelDHL and RDFox (see Figure~\ref{fig:diffdepths}(left)).
The downtrend of reasoning times also exists
when handling the datasets nptu-$4m$-$i$, $i\in\{1,2,4,8,16\}$ (see Figure~\ref{fig:diffdepths}(right)).
Although the PTU and NPTU datasets are not real-world examples, these experiments indeed
verify that MG-depths determines the reasoning time in parallel
considering that the input ontology sizes are close.

@@

We now discuss the issue of parallel tractability based on the experimental
results of ptu-$4m$-1 and nptu-$4m$-1.
For the dataset ptu-$4m$-1, the trends of reasoning times are different
between ParallelDHL and RDFox.
Table~\ref{tab:speedup} shows that
RDFox does not benefit from the allocation of several threads for ptu-$4m$-1;
the speedups under different numbers of threads stay around 1.
This means that reasoning times are not reduced with an increasing number of threads.
ParallelDHL shows a better speedup effect. With only one allocated thread, ParallelDHL finishes
the materialization of ptu-$4m$-1 in $135.79$ seconds (Table~\ref{tab:speedup}). When four and more threads
are allocated, the reasoning performance is improved significantly.
The maximal speedup is $5.23$ using $24$ threads.
Specially, with $16$ or more threads, ParallelDHL
requires less time to finish the materialization compared to RDFox.
From Figure~\ref{fig:diffthreads}(left), there is an obvious downtrend of the reasoning times
from one thread to $24$ threads.
The main reason for the difference between ParallelDHL and RDFox
is that the PTU ontology is tractable in parallel and ParallelDHL
is optimized based on SWD paths. The PTU ontology follows the
simple-concept restriction. Thus, there is no twisting path in
the materialization graph of ptu-$4m$-1 although all articles are involved
in one reference chain. This allows for computing the relation $S_{\textit{\!\tiny
    rch}}$ only once in the second iteration of \ref{alg3:updateG} of Algorithm~$\mathsf{A}_{\text{prc}}$
when handling ptu-$4m$-1 (referring to the analysis for Example~\ref{exp:simpleC}).
It can also be checked that all the axioms of the form \texttt{CollegeArticle}$(a_i)$ occur in
an SWD path. Thus, the optimization based on SWD paths works when handling ptu-$4m$-1.
For RDFox, since it is not optimized specially for this case, it performs the task of
materialization over ptu-$4m$-1 similarly to that over nptu-$4m$-1;
specifically, the axioms of the form \texttt{CollegeArticle}$(a_i)$ are derived sequentially.
From Figure~\ref{fig:diffdepths}(left), we can see that the optimization used in ParallelDHL
also leads to a better performance compared to
RDFox when handling ptu-$4m$-1, ptu-$4m$-2 and ptu-$4m$-4 using $24$ threads.

### 2018-02-27 parallel-tractability-J-v9/introduction.tex
final polishing

According to \citet{MotikNPHO14}, many real large-scale ontologies are essentially expressed in ontology languages
that can be rewritten into datalog rules. We follow this argumentation and focus on such datalog rewritable ontology languages in this paper.
The main target of this paper is to identify classes of datalog rewritable ontologies such that materialization
over these ontologies is tractable in parallel, i.e., falls into the parallel complexity class \texttt{NC} \cite{Raymond95}.
This complexity class consists of problems that can be solved efficiently in parallel.
To show that a problem is in \texttt{NC}, one can give an \emph{\texttt{NC} algorithm} that handles this problem
using parallel computation \cite{Raymond95}.
An \texttt{NC} algorithm is required to terminate in parallel poly-logarithmic time.
However, current materialization algorithms of datalog rewritable ontology languages
(e.g., the core algorithm used in RDFox \cite{MotikNPHO14}) are not \texttt{NC} algorithms
since they are designed for general datalog programs and have \texttt{PTime}-complete complexity.
Thus, we first give \texttt{NC} algorithms that perform materialization
and then identify the corresponding classes of datalog rewritable ontologies (called \emph{parallel tractability classes})
that can be handled by these \texttt{NC} algorithms. To make the proposed \texttt{NC} algorithms practical,
we also discuss how to optimize and implement them.

In the practical part of this work, we study specific datalog rewritable ontology languages.
We first focus on the ontology language DL-Lite \cite{CalvaneseGLLR07} to clarify how to apply the above \texttt{NC} algorithms
to study parallel tractability of ontology materialization.
We show that DL-Lite$_{\text{core}}$ and DL-Lite$_\mathcal{R}$ are tractable in parallel.
We then study the ontology language Description Horn Logic (DHL) \cite{GrosofHVD03}, which is the intersection of datalog and OWL
in terms of expressivity. We give a case of a DHL ontology where materialization can hardly be parallelized.
Based on the analysis of this case, we propose to restrict the usage of DHL such that materialization over the
restricted ontologies can be handled by the proposed \texttt{NC} algorithms.
We further extend the results to an extension of DHL that also allows complex role inclusion axioms.
Finally, we analyze the well-known benchmark, LUBM, and the real-world dataset, YAGO, and
show that these ontologies following the
proposed restrictions belong to the parallel tractability classes.
We implement a system based on an optimized \texttt{NC} algorithm for DHL materialization,
which we compare to the state-of-the-art reasoner RDFox.
The experimental results show that the optimizations proposed in this paper result in a
better performance on ontologies that are
tractable in parallel compared with RDFox.

The remainder of the paper is organized as follows. In Section~\ref{sec:background}, we introduce some basic notions.
We then give two \texttt{NC} algorithms for ontology materialization in Section~\ref{sec:ptclass}.
We study parallel tractability of materialization in DL-Lite, DHL and an extension of DHL in Section~\ref{sec:ptonto}, respectively.
In Section~\ref{sec:practicalAlg},
we discuss how to optimize and implement the given \texttt{NC} algorithms.
We analyze the LUBM and the YAGO ontologies and evaluate our implementation in Section~\ref{sec:evaluation}.
Finally, we discuss related work in Section~\ref{sec:related} and conclude in Section~\ref{sec:conclusion}. Detailed proofs of the theorems and lemmas in this paper are given in the appendix.

>>>>

According to \citet{MotikNPHO14}, many real large-scale ontologies are essentially expressed in ontology languages
that can be rewritten into datalog rules. We follow this argumentation and focus on such datalog rewritable ontology languages in this paper.
The main target of this paper is to identify classes of datalog rewritable ontologies such that materialization
over these ontologies is tractable in parallel, i.e., falls into the parallel complexity class \texttt{NC} \cite{Raymond95}.
This complexity class consists of problems that can be solved efficiently in parallel.
To show that a problem is in \texttt{NC}, one can give an \emph{\texttt{NC} algorithm} that handles this problem
using parallel computation \cite{Raymond95}.
An \texttt{NC} algorithm is required to terminate in parallel poly-logarithmic time.
However, \mbox{current} materialization algorithms of datalog rewritable ontology languages
(e.g., the core algorithm used in RDFox \cite{MotikNPHO14}) are not \texttt{NC} algorithms
since they are designed for general datalog programs and have \texttt{PTime}-complete complexity.
Thus, we first give \texttt{NC} algorithms that perform materialization
and then identify the corresponding classes of datalog rewritable ontologies (called \emph{parallel tractability classes})
that can be handled by these \texttt{NC} algorithms. To make the proposed \texttt{NC} algorithms practical,
we also discuss how to optimize and implement them.

In the practical part of this work, we study specific datalog rewritable ontology languages.
We first focus on the ontology language DL-Lite \cite{CalvaneseGLLR07} to clarify how to apply the above \texttt{NC} algorithms
to study parallel tractability of ontology materialization.
We show that DL-Lite$_{\text{core}}$ and DL-Lite$_\mathcal{R}$ are tractable in parallel.
We then study the ontology language Description Horn Logic (DHL) \cite{GrosofHVD03}, which is the intersection of datalog and OWL
in terms of expressivity. We give a case of a DHL ontology where materialization can hardly be parallelized.
Based on the analysis of this case, we propose to restrict the usage of DHL such that materialization over the
restricted ontologies can be handled by the proposed \texttt{NC} algorithms.
We further extend the results to an extension of DHL that also allows complex role inclusion axioms.
Finally, we analyze the well-known benchmark, LUBM, and the real-world dataset, YAGO, and
show that these ontologies satisfy the
proposed restrictions and, hence, belong to the parallel tractability classes.
We implement a system based on an optimized \texttt{NC} algorithm for DHL materialization,
which we compare to the state-of-the-art reasoner RDFox \cite{MotikNPHO14}.
The experimental results show that the optimizations proposed in this paper result in a
better performance on ontologies that are
tractable in parallel compared to RDFox.

The remainder of the paper is organized as follows. In Section~\ref{sec:background}, we introduce some basic notions.
We then give two \texttt{NC} algorithms for ontology materialization in Section~\ref{sec:ptclass}.
We study parallel tractability of materialization in DL-Lite, DHL and an extension of DHL in Section~\ref{sec:ptonto}.
In Section~\ref{sec:practicalAlg},
we discuss how to optimize and implement the given \texttt{NC} algorithms.
We analyze the LUBM and the YAGO ontologies and evaluate our implementation in Section~\ref{sec:evaluation}.
Finally, we discuss related work in Section~\ref{sec:related} and conclude in Section~\ref{sec:conclusion}. Detailed proofs of the theorems and lemmas in this paper are given in the appendix.

### 2018-02-27 parallel-tractability-J-v9/parallelTractabilityDatalog.tex
final polishing

\begin{proof}[Proof sketch] This lemma can be proved by performing
an induction on $T_R^{\omega}(\textbf{I})$.
The stage (see the related contents in Section~\ref{sec:background}) of $P$
is the lower-bound of the depth of the materialization graphs. Based on the previous induction,
one can further check that, for the materialization graph $\mathcal{G}$ constructed by Algorithm~$\mathsf{A}_{\text{bsc}}$,
its depth equals the depth of the stage.
\end{proof}

>>>>

\begin{proof}[Proof sketch] This lemma can be proved by performing
an induction on $T_R^{\omega}(\textbf{I})$.
The stage (see the related contents in Section~\ref{sec:background}) of $P$
is the lower-bound of the depth of the materialization graphs. Based on the previous induction,
one can further check that, for the materialization graph $\mathcal{G}$ constructed by Algorithm~$\mathsf{A}_{\text{bsc}}$,
its depth is equal to the depth of the stage.
\end{proof}

### 2018-02-27 parallel-tractability-J-v9/parallelTractabilityOntMat.tex
final polishing

The original rule set for RDFS reasoning is
given by \citet{RDFSrec04}. By applying the rules in this rule set,
new schema data could be derived. Thus, the reasoning task of RDFS
is different from the task of materialization. Further, an RDFS ontology allows
statements about blank nodes which act like variables. This possibly
leads to infinite entailments. Additionally, the original rule set is
incomplete with the RDF restriction that
blank nodes cannot occur in predicate positions \cite{Horst05},
however, by allowing blank nodes in predicate position
the rule set turns out to be complete. The computational complexity
of RDFS reasoning is NC-complete \cite{Horst05},
which is not considered to be tractable in parallel.

@@

In this work, we study the task of materialization, hence, we can only focus on
newly entailed instance data for the RDFS ontologies. To this end,
we assume that any class of RDFS ontologies has the same schema data (\uppercase\expandafter{\romannumeral1}). In this way,
axioms such as $\beta$ cannot contribute to the computational complexity, since they
only apply to schema data. Further, we assume blank nodes not to occur in
any class of RDFS ontologies
(\uppercase\expandafter{\romannumeral2}). This allows for expressing RDFS ontology
in DHL (see Table~\ref{tab:rdfs}). Based on the two restrictions (\uppercase\expandafter{\romannumeral1})
and (\uppercase\expandafter{\romannumeral2}), the materialization of RDFS ontologies
is tractable in parallel (see Section~\ref{sec:DHL}).

>>>>

The original rule set for RDFS reasoning is
given by \citet{RDFSrec04}. By applying the rules in this rule set,
new schema data could be derived. Thus, the reasoning task of RDFS
is different from the task of materialization. Further, an RDFS ontology allows
statements about blank nodes which act like variables. This possibly
leads to infinite entailments. Additionally, the original rule set is
incomplete with the RDF restriction that
blank nodes cannot occur in predicate positions \cite{Horst05},
however, by allowing blank nodes in predicate position
the rule set turns out to be complete. The computational complexity
of RDFS reasoning is \texttt{NC}-complete \cite{Horst05},
which is not considered to be tractable in parallel.

@@

In this work, we study the task of materialization, hence, we can only focus on
newly entailed instance data for the RDFS ontologies. To this end,
we assume that any class of RDFS ontologies has the same schema data (\uppercase\expandafter{\romannumeral1}). In this way,
axioms such as $\beta$ cannot contribute to the computational complexity, since they
only apply to schema data. Further, we assume blank nodes not to occur in
any class of RDFS ontologies
(\uppercase\expandafter{\romannumeral2}). This allows for expressing RDFS ontologies
in DHL (see Table~\ref{tab:rdfs}). Based on the two restrictions (\uppercase\expandafter{\romannumeral1})
and (\uppercase\expandafter{\romannumeral2}), the materialization of RDFS ontologies
is tractable in parallel (see Section~\ref{sec:DHL}).
